%!TEX root = ../../da.tex
\cleardoublepage

\thispagestyle{plain}

{\normalfont\Large\noindent Acknowledgements}
\vspace{\baselineskip}

\noindent
Like most things worth doing, this thesis was not a solitary effort.

First of all, I would like to thank Matthias Rupp, who replied to my first email asking about quantum mechanics and machine learning and has not stopped answering since.
Equally big thanks belong to Matthias Scheffler and Klaus-Robert Müller, who welcomed me into their groups and supported and guided me in my work. I am very grateful for the opportunity to work in such excellent environments, and to benefit so greatly from their resources and opportunities. Particular thanks is also due to the Institute for Pure and Applied Mathematics, where many main ideas in this thesis were first developed.

This work would have been impossible without my co-authors, co-workers, and colleagues. The first is Florian Knoop, who taught me almost everything I know about thermal transport, and who never turned down a \scare{small beer}. Thanks also to Christian Carbogno, who taught me the rest. In no particular order, I would like to thank other colleagues who have answered my questions, indulged mad ideas, and have otherwise contributed to an overall fantastic working environment: 
Mihail Bogojeski,
Stefan Chmiela,
Lucas Deecke,
Thorben Frank,
Adrian Hill,
Sebastian Kokott,
Hagen-Henrik Kowalski,
Marcin Krynski,
Yair Litman,
Dmitrii Maksimov (and friends),
Niklas Schmitz,
Daniel Speckhart,
Chris Sutton,
Thomas Purcell,
the regulars of the \ase bar,
and many more who would not fit on the page.

Many thanks are also due to administrative staff, in particular Julia Pach, Hanna Krauter, Rayya Douedari, Andrea Gerdes, and Jana Peich, who have been an invaluable help in navigating bureaucracy and getting things done, often at the very last minute.

Outside of the office, I thank my friends in Berlin, and all the other places, physical and virtual, for making these last few years the best ones yet, and for making this city my home. Without your support, and a reminder of life outside work, this would have been impossible.

At this final chapter of my formal education, I would like to thank those that started it: Annegret Rehmet, who first taught me about learning, Peter Meggle, who showed me that computers and mathematics go hand in hand, and Benjamin Durst, who restored my faith in physics, and many others who helped along the way.

I would also like to acknowledge Keysound Recordings for providing music at a critical moment, and Sven Pfleiderer for donating a spare laptop at the perfect time.
\\\\
Last, but very much not least, I would like to thank my parents, Heiderose Langer and Matthias Bier, without whom none of this would have been possible. Thank you for everything.
\\\\
\hspace*{\fill} Berlin,\\
\hspace*{\fill} 5th June 2023
